---
title: "Supplementing a Metric Repository"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Supplementing a Metric Repository}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(val.criterion)
library(val.meter)
```

A common question from organizations evaluating `val.criterion` is whether they
can mix in metrics of their *own* -- for example, an internal notion of a
"reputable" maintainer, or a bespoke security score -- alongside the metrics
published by the R Validation Hub. They can, and this vignette walks through the
three moving parts:

1. **registering** a new metric so `val.meter` knows its name and type,
2. **publishing** its values in a supplementary `PACKAGES` file, and
3. **filtering** on a mix of criteria drawn from two separate metric
   repositories.

Throughout, we simulate everything locally with `val.meter::random_repo()` so
the example runs offline.

## Registering a new metric

Metrics live in `val.meter`. Filtering by a metric requires that the metric be
*registered*, because `val.criterion` uses the metric's declared type to
interpret the character values stored in a `PACKAGES` file (every value in a
repository index is stored as text). An unregistered field has no known type and
cannot be coerced reliably.

We register metrics with `val.meter::impl_data(metric = TRUE)`. The
[*Adding a Metric*](https://pharmar.github.io/val.meter/articles/adding-metric.html)
vignette in `val.meter` covers this in depth; here we add a small
`reputable_maintainer` metric that flags packages for which the *R Validation
Hub* is a copyright holder.

```{r}
impl_data(
  "reputable_maintainer",
  title = "Reputable Maintainer",
  metric = TRUE,
  class = S7::class_logical,
  function(pkg, resource, field, ...) {
    authors <- pkg$desc$get_authors()
    any(vapply(
      authors,
      function(a) "cph" %in% a$role &&
        grepl("R Validation Hub", paste(a$given, a$family)),
      logical(1)
    ))
  }
)
```

Once registered, the metric shows up alongside the built-in metrics:

```{r}
"reputable_maintainer" %in% names(metrics())
```

## Simulating an upstream metric repository

A metric repository is an ordinary CRAN-style repository whose
`src/contrib/PACKAGES` file is tagged `Format: Metrics` and carries one
`Metric/<field>@R:` entry per metric. `val.meter::random_repo()` writes one for
us -- standing in for the metrics an organization might consume from the R
Validation Hub. We'll call it `valor`.

```{r, warning = FALSE}
set.seed(1)
valor <- random_repo(n = 8)
```

Reading its index shows the metric columns we can filter on:

```{r}
valor_url <- file.path(sub("^file://", "", valor), "src", "contrib", "PACKAGES")
base_db <- read.dcf(valor_url)
base_db <- base_db[!is.na(base_db[, "Package"]), , drop = FALSE]

grep("^Metric/", colnames(base_db), value = TRUE)
```

## Building a supplementary `PACKAGES` file

Now suppose "acme" -- our organization -- wants to contribute its own
`reputable_maintainer` assessment without touching the upstream repository. We
publish a *second* metric repository containing only our metric.

The one rule that makes the two repositories joinable is that each record must
reuse the upstream `Package`, `Version` and `MD5sum` -- those three fields are
the join key. A supplementary repository omits the package metrics it does not
provide and lists only its own.

```{r}
acme_dir <- tempfile("acme")
dir.create(file.path(acme_dir, "src", "contrib"), recursive = TRUE)

set.seed(42)
records <- c("Format: Metrics", "")
for (i in seq_len(nrow(base_db))) {
  records <- c(
    records,
    paste0("Package: ", base_db[i, "Package"]),
    paste0("Version: ", base_db[i, "Version"]),
    paste0("MD5sum: ",  base_db[i, "MD5sum"]),
    # simulate acme's assessment for the demo
    paste0("Metric/reputable_maintainer@R: ", sample(c("TRUE", "FALSE"), 1)),
    ""
  )
}
writeLines(records, file.path(acme_dir, "src", "contrib", "PACKAGES"))
acme <- paste0("file://", acme_dir)
```

The result is a minimal, valid metric repository:

```{r}
cat(readLines(file.path(acme_dir, "src", "contrib", "PACKAGES"), n = 7), sep = "\n")
```

## Filtering on metrics from two repositories

Each metric repository reduces to an `available.packages()`-style matrix. A
small helper reads one, keeping only the metric columns:

```{r}
read_metrics <- function(repo) {
  url <- file.path(sub("^file://", "", repo), "src", "contrib", "PACKAGES")
  fields <- grep("^Metric/", colnames(read.dcf(url)), value = TRUE)
  db <- available.packages(repos = repo, fields = fields, filters = list())
  db[!is.na(db[, "Package"]), , drop = FALSE]
}
```

To filter on metrics from both repositories at once, join them on the
`Package`/`Version`/`MD5sum` key into a single metric matrix. This gives every
package a `downloads_total` column (from `valor`) *and* a
`reputable_maintainer` column (from `acme`):

```{r}
combined <- merge(
  read_metrics(valor),
  read_metrics(acme),
  by = c("Package", "Version", "MD5sum"),
  all = TRUE
)

grep("downloads_total|reputable_maintainer", colnames(combined), value = TRUE)
```

`package_filter()` accepts a `db` argument -- a function returning the metric
matrix to filter against -- so we hand it our combined matrix. The filter
expression then references metrics from either repository by name, as flat
variables:

```{r}
options(
  repos = valor,
  available_packages_filters = package_filter(
    {
      # passes R CMD check *and* is maintained by a reputable org
      r_cmd_check_error_count == 0 & reputable_maintainer
    },
    db = function() as.matrix(combined)
  )
)
```

With the filter installed, base R's `available.packages()` transparently drops
packages that fail the policy -- their `Repository` is rewritten to
`"<filtered>"`:

```{r}
pkgs <- available.packages()
pkgs[, c("Package", "Repository")]
```

`install.packages()` honours the same filter, so a package excluded by policy
cannot be installed through the normal workflow.

## A note on joining repositories

`val.criterion` reads the repositories in `options("val.criterion.repos")` by
row-binding their indices, which does not by itself *column-join* two
repositories that describe the same packages with different metrics. Until
native joining is available, supply the joined matrix explicitly through the
`db` argument, as shown above. The join key is always
`c("Package", "Version", "MD5sum")`.
